A generalization of the Bradley-Terry model for draws in chess

### 2020-09-16 inst/jebo_chess_rebuttal.txt

I am in total agreement with this observation.  My thinking on this
issue has been strongly affected by my recent work applying
Bradley-Terry to Formula 1 motor racing.  In F1, drivers generally
start their careers as weak competitors, become stronger with
experience, and finally become uncompetitive due to age [although
frankly Lewis Hamilton at 35yo seems to buck this general trend, as
recent events in Tuscany would indicate].  I would suggest that chess
players have a similar career arc, although Formula 1 Teams typically
take a dim view of underperforming drivers---whom they rapidly and
unceremonially drop.  The career arc for these competitors would have
a truncated right hand side, unlike chess players who typically
decline gradually over decades [although I would observe that
Kasparov's current FIDE rating of 2812 is not that much less than his
peak of 2851].

Specifically, the dataset in question stretches from 1981 to 2017, a
very wide range.  Generalized Bradley-Terry, together with reified
entities, is capable of handling strengths varying over a career: one
would introduce a "mid-career monster", or perhaps an "experience
monster".  I would add here that monsters' strength is inherently
positive, so one could deal with waning strength with age by
introducing a "youth monster" who helps younger players.  But such
analysis is, I would suggest, firmly in the category of further work.
I have added a discussion of this issue to the manuscript (RED), in
two parts: firstly when introducing the dataset, and secondly under
Further Work.

> 2) The author neglects to motivate the paper beyond its application
> within the chess world.  The tournament literature's applications lie
> primarily, though definitely not exclusively, within compensation and
> labor.  How could this extension be used within that context?  Where
> might economists see the first-mover advantage in action? Where might
> the extension to allow draws be especially lucrative? Etc.
>

(manuscript changes addressing this comment are marked in ORANGE).

The original submission discussed applications to MasterChef
Australia, consumer behaviour and mating preference, but I was not
aware of tournament literature as a distinct academic field.  I have
been studying Lazear and Rosen and pondering how reified monsters
could help understanding tournaments in a workplace remuneration
context.  Lazear and Rosen supply a good rationale to consider only
rank in situations such as Formula 1 motor racing

First-mover advantage is closely related to incumbent advantage,
something which had not occurred to me until now.  One of the papers
cited [Hankin 2010] uses a variant of Bradley-Terry to assess
incumbent advantage in biology (territory defense by lizards) and I
have added a more explicit pointer to this.

Returning to Lazear and Rosen, schemes that reward rank are common in
sports (notably Formula 1, discussed above, and sculling, both good
applications of the Plackett-Luce extension of Bradley-Terry).

Also, I have discussed Formula 1 motor racing above under part 1.

### 2020-09-16 inst/jebo_chess_revision2.tex
work on incumbent advantage

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).

@@

>>>>

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).  \textcolor{orange}{First-mover advantage is closely
  related to incumbent advantage, and the techniques presented here
  have been applied to ecological incumbency~\cite{hankin2010}.}

@@

\textcolor{orange}{The maximum likelihood estimate for strengths
  furnishes a natural ranking for the competitors.  \cite{lazear1981}
  observe that such ranking systems occur naturally in organizational
  situations (tournament theory) and the analysis presented here might
  be applicable to this situation, with the advantage of allowing for
  both draws and incumbent advantage.}
